\documentclass[12pt, leqno]{article} %% use to set typesize
\include{common}

\begin{document}

\hdr{HW 5}

You may (and should) talk about problems with each other and with me,
providing attribution for any good ideas you might get.  Your final
write-up should be your own.

\paragraph*{1: Minimize this!}
Consider $A \in \bbR^{n \times n}, b \in \bbR^n$ with $A$ symmetric
and $b$ not an eigenvector, and define
\[
  \phi(x) = \frac{1}{2} x^T A x - x^T b.
\]
We wish to minimize $\phi$ subject to the constraint $x^T x = 1$,
via the Lagrangian
\[
  L(x,\mu) = \frac{1}{2} x^T A x - x^T b - \frac{\mu}{2} (x^T x - 1).
\]
\begin{enumerate}
\item Express $x$ at a stationary point in terms
  of $A$, $b$, and $\mu$.
\item Argue that the condition $\|x\|^2 = 1$, given the expression
  from the previous step, implies
  singularity of the matrix
  \[
    \begin{bmatrix} (A-\mu I)^2 & b \\ b^T & 1 \end{bmatrix}
  \]
\item
  Eliminating the $z$ variable in
  \[
    \begin{bmatrix}
      (A-\mu I)^2 & b \\
      b^T & 1
    \end{bmatrix}
    \begin{bmatrix} y \\ z \end{bmatrix} = 0,
  \]
  show that $\mu$ satsifies the quadratic eigenvalue problem
  \[
    \left[ (A^2-bb^T) - 2 \mu A + \mu^2 I \right] y = 0.
  \]
  Solving via {\tt polyeig} gives
  us all possible $\mu$ in $O(n^3)$ time.
\item
  At the constrained minimizer, we must satisfy that
  $v^T (A-\mu I) v$ is positive for all $v$ s.t.~$v^T x = 0$.
  Argue that this implies $\mu < \lambda_2(A)$.
\item
  Following the divide-and-conquer idea and the argument
  from the previous step, argue that $\phi$
  has no more than three constrained minimizers.
\end{enumerate}

\paragraph*{2: Catching cosines}
Suppose $A = H+S \in \bbR^{n \times n}$ where $H$ is symmetric
and $S$ is skew symmetric.  Let $A = U \Sigma V^T$ be the SVD,
and let $\theta_k$ be the angle between the $k$th left and right
singular vectors.  Show that
\[
  \frac{\lambda_{\min}(H)}{\sigma_k} \leq \cos(\theta_k) \leq \frac{\lambda_{\max}(H)}{\sigma_k}
\]
where $\lambda_{\min}(H)$ and $\lambda_{\max}(H)$ are the smallest and largest
eigenvalues of $H$.

\paragraph*{3: Think positive}
Consider the Jacobi iteration for an elementwise non-negative
matrix $A$ and a non-negative
right hand side vector $b$.  Show that if the initial guess
$x^{(0)}$ is elementwise greater than $x$, then $x^{(k)} \geq x$
elementwise for any even $k$ and $x^{(k)} \leq x$ elementwise for
any odd $k$.

\paragraph*{4: Smooth operator}
Write a code to implement the Gauss-Seidel algorithm for the
one-dimensional model problem (a simple tridiagonal) with a random right
hand side.  At each step, compute the error and plot the absolute value
of the FFT of the error.  What do you notice about the behavior with
increasing numbers of sweeps?  Your submission should include a plot
that illustrates the point.

\end{document}
