
\section{Logistics}

CS 6210 is a graduate-level introduction to numerical linear algebra. We
will study direct and iterative methods for linear systems, least
squares problems, eigenvalue problems, and singular value problems.
There is a detailed syllabus and rough schedule at the class web:
\begin{center}
  \url{http://www.cs.cornell.edu/~bindel/class/cs6210-f16/}
\end{center}
The web page is also your source for the homework, lecture notes,
and course announcements.  The web page also points to a discussion
forum (Piazza) and the Course Management System (CMS) used for homework
submissions.  There is also a link to a GitHub repository with
the source for these notes.  If you find an error, feel free to
help me fix it!

I will assume that you have already had a course in linear algebra and
that you know how to program.  A previous numerical methods course  is
not required.  The course will involve MATLAB programming, so it will
also be helpful -- but not strictly necessary -- for you to have prior
exposure to MATLAB.  If you do not know MATLAB already but you have
prior programming experience, you can learn MATLAB quickly enough.
Octave is a viable substitute for MATLAB for the work in this course.

Several of you have already taken CS 4220 (with me).  You will find that
some sections of the course are review, particularly at the beginning.
However, because CS 6210 is a graduate course with more assumed
background, and because we will spend the entire semester on numerical
linear algebra, we will cover a wider variety of topics in numerical
linear algebra and at a deeper level.  If you have concerns about
whether this is the class for you, please come and talk to me.

This course has proven surprisingly popular, and we hit the room cap
during pre-enrollment.  We have since moved to a larger room that can
accomodate 50 (rather than 40), and I hope to be able to seat all comers.
If you {\em are not} planning to take the course for credit, please
help us out by releasing your seat.  If you {\em are} planning to take
the course for credit, please send me email with your name and NetID,
and I will try to make sure something sane works out.
