\section{Problems to ponder}

\begin{enumerate}
\item
  How do we accurately evaluate $\sqrt{1+x}-\sqrt{1-x}$ when $x \ll 1$?
\item
  How do we accurately evaluate $\ln \sqrt{x+1}-\ln\sqrt{x}$ when $x \gg 1$?
\item
  How do we accurately evaluate $(1-\cos(x))/\sin(x)$ when $x \ll 1$?
\item
  How would we compute $\cos(x)-1$ accurately when $x \ll 1$?
\item
  The {\em Lamb-Oseen vortex} is a solution to the 2D Navier-Stokes
  equation that plays a key role in some methods for computational
  fluid dynamics.  It has the form
  \[
    v_{\theta}(r,t) = \frac{\Gamma}{2 \pi r}
      \left( 1 - \exp\left( \frac{-r^2}{4 \nu t} \right) \right)
  \]
  How would one evaluate $v(r,t)$ to high relative accuracy for all
  values of $r$ and $t$ (barring overflow or underflow)?
\item
  For $x > 1$, the equation $x = \cosh(y)$ can be solved as
  \[
    y = -\ln\left(x - \sqrt{x^2-1}\right).
  \]
  What happens when $x = 10^8$?  Can we fix it?
\item
  The difference equation
  \[
    x_{k+1} = 2.25x_k - 0.5 x_{k-1}
  \]
  with starting values
  \[
    x_1 = \frac{1}{3}, \qquad x_2 = \frac{1}{12}
  \]
  has solution
  \[
    x_k = \frac{4^{1-k}}{3}.
  \]
  Is this what you actually see if you compute?  What goes wrong?
\item
  Considering the following two MATLAB fragments:
\begin{lstlisting}
    % Version 1
    f = (exp(x)-1)/x;

    % Version 2
    y = exp(x);
    f = (1-y)/log(y);
\end{lstlisting}
  In exact arithmetic, the two fragments are equivalent.  In floating
  point, the first formulation is inaccurate for $x \ll 1$, while the
  second formulation remains accurate.  Why?
\item
  Running the recurrence $E_n = 1-nE_{n-1}$ {\em forward} is an
  unstable way to compute $\int_0^1 x^n e^{x-1} \, dx$.  However, we
  can get good results by running the recurrence {\em backward} from
  the estimate $E_n \approx 1/(N+1)$ starting at large enough $N$.
  Explain why.  How large must $N$ be to compute $E_{20}$ to near
  machine precision?
\item
  How might you accurately compute this function for $|x| < 1$?
  \[
    f(x) = \sum_{j=0}^\infty \left( \cos(x^j)-1 \right)
  \]
\end{enumerate}
